---
title: "Coordinate Mapping in gazeneuro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coordinate Mapping in gazeneuro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gazeneuro)
library(tidyverse)
```

## Overview

The coordinate mapping functionality in `gazeneuro` allows you to convert gaze tracking coordinates to anatomical locations in neuroimaging space. This is essential for understanding which brain regions participants were looking at during image viewing.

## Coordinate Systems

The package handles three coordinate systems:

1. **Fractional coordinates (0-1)**: Normalized coordinates from gaze tracking
2. **Voxel coordinates**: Integer indices into the 3D image array
3. **World/MM coordinates**: Physical locations in millimeters

## Basic Usage

### Single Point Mapping

To map a single gaze point to anatomical coordinates:

```{r, eval=FALSE}
# Load your data
nifti_data <- preload_nifti_data("brain.nii.gz")

# Map a single point at the center of the display
location <- locate_single_point(
  x = 0.5,      # Center X (0-1)
  y = 0.5,      # Center Y (0-1)
  z = 0.5,      # Middle slice (0-1)
  nifti_data = nifti_data,
  plane = "AXIAL"
)

# Access the coordinates
print(location$mm)   # World coordinates in mm
print(location$vox)  # Voxel indices
print(location$values[[1]]$value)  # Intensity value
```

### Batch Processing

For processing all gaze points from an eye tracking session:

```{r, eval=FALSE}
# Integrate gaze and slice data
integrated <- integrate_all_gaze_points(gaze_data, z_axis_data)

# Map all points to anatomical locations
locations <- map_gaze_to_anatomy(integrated, nifti_data)

# View the first few rows
head(locations)
```

## Analyzing Gaze Patterns in Anatomical Space

### Find Most-Viewed Voxels

```{r, eval=FALSE}
# Get summary by unique voxel locations
voxel_summary <- batch_locate_gaze(
  integrated, 
  nifti_data, 
  summarize = TRUE
)

# Top 10 most viewed voxels
top_voxels <- voxel_summary %>%
  arrange(desc(n_gazes)) %>%
  head(10)

print(top_voxels)
```

### Visualize Anatomical Locations

```{r, eval=FALSE}
# Create visualization of coordinate mappings
plot_coordinate_mapping(locations, nifti_data)
```

## Working with Different Imaging Planes

The coordinate mapping works across all three standard imaging planes:

```{r, eval=FALSE}
# Process SAGITTAL plane
sagittal_gazes <- integrated %>%
  filter(plane == "SAGITTAL")

sagittal_locations <- map_gaze_to_anatomy(
  sagittal_gazes, 
  nifti_data
)

# Process CORONAL plane
coronal_gazes <- integrated %>%
  filter(plane == "CORONAL")

coronal_locations <- map_gaze_to_anatomy(
  coronal_gazes, 
  nifti_data
)
```

## Extracting Regional Statistics

### By Anatomical Region

If you have a brain atlas, you can aggregate gaze statistics by region:

```{r, eval=FALSE}
# Assuming you have an atlas aligned to your image
atlas_data <- preload_nifti_data("atlas.nii.gz")

# Get atlas labels for each gaze location
locations_with_regions <- locations %>%
  mutate(
    # Get region ID from atlas
    region_id = sapply(1:n(), function(i) {
      safe_get_value(
        atlas_data$data, 
        c(vox_x[i], vox_y[i], vox_z[i]),
        atlas_data$dims
      )
    })
  )

# Summarize by region
regional_summary <- locations_with_regions %>%
  filter(!is.na(region_id), region_id > 0) %>%
  group_by(region_id) %>%
  summarise(
    n_gazes = n(),
    total_duration = max(time_sec) - min(time_sec),
    mean_intensity = mean(intensity, na.rm = TRUE)
  )
```

### Heatmap Generation

Create a 3D heatmap of gaze density:

```{r, eval=FALSE}
# Initialize empty heatmap
heatmap_3d <- array(0, dim = nifti_data$dims)

# Accumulate gaze counts
for (i in 1:nrow(locations)) {
  vox <- c(locations$vox_x[i], locations$vox_y[i], locations$vox_z[i])
  
  # Check bounds
  if (all(vox >= 0) && all(vox < nifti_data$dims)) {
    heatmap_3d[vox[1]+1, vox[2]+1, vox[3]+1] <- 
      heatmap_3d[vox[1]+1, vox[2]+1, vox[3]+1] + 1
  }
}

# Smooth the heatmap (optional)
# heatmap_3d <- smooth_3d(heatmap_3d)

# Save as NIfTI
# write_nifti(heatmap_3d, "gaze_heatmap.nii.gz")
```

## Advanced Usage

### Custom Coordinate Transformations

For specialized analyses, you can work directly with the coordinate transformation functions:

```{r, eval=FALSE}
# Get the NVImage object
nvimg <- nifti_data$nvimage

# Custom fractional coordinate
frac_coord <- c(0.75, 0.25, 0.5)

# Convert to mm
mm_coord <- nvimg$convertFrac2MM(frac_coord)
print(paste("MM coordinates:", paste(mm_coord[1:3], collapse=", ")))

# Convert to voxel
vox_coord <- nvimg$mm2vox(mm_coord[1:3])
print(paste("Voxel indices:", paste(vox_coord, collapse=", ")))
```

### Time-Based Analysis

Analyze how gaze patterns change over time:

```{r, eval=FALSE}
# Divide session into time bins
locations_with_bins <- locations %>%
  mutate(
    time_bin = cut(time_sec, breaks = 10, labels = FALSE)
  )

# Analyze spatial spread over time
time_analysis <- locations_with_bins %>%
  group_by(time_bin) %>%
  summarise(
    n_gazes = n(),
    spatial_spread_mm = sqrt(var(mm_x) + var(mm_y) + var(mm_z)),
    mean_x_mm = mean(mm_x),
    mean_y_mm = mean(mm_y),
    mean_z_mm = mean(mm_z)
  )

# Plot spatial spread over time
ggplot(time_analysis, aes(x = time_bin, y = spatial_spread_mm)) +
  geom_line() +
  geom_point() +
  labs(x = "Time Bin", y = "Spatial Spread (mm)",
       title = "Gaze Pattern Dispersion Over Time")
```

## Performance Considerations

For large datasets:

1. **Process in chunks**: Don't load all gaze data at once
2. **Use parallel processing**: For batch operations
3. **Pre-filter data**: Remove invalid gaze points before mapping

```{r, eval=FALSE}
# Example of chunk processing
chunk_size <- 10000
n_chunks <- ceiling(nrow(integrated) / chunk_size)

all_locations <- list()

for (i in 1:n_chunks) {
  start_idx <- (i - 1) * chunk_size + 1
  end_idx <- min(i * chunk_size, nrow(integrated))
  
  chunk_locations <- map_gaze_to_anatomy(
    integrated[start_idx:end_idx, ],
    nifti_data
  )
  
  all_locations[[i]] <- chunk_locations
}

# Combine results
final_locations <- do.call(rbind, all_locations)
```

## Integration with Other Packages

The coordinate mapping output can be used with other neuroimaging packages:

```{r, eval=FALSE}
# Export for use with FSL or SPM
write.csv(locations, "gaze_coordinates.csv", row.names = FALSE)

# Create a point cloud for visualization
library(rgl)
plot3d(locations$mm_x, locations$mm_y, locations$mm_z,
       col = heat.colors(100)[cut(locations$intensity, 100)],
       size = 3)
```
